---
title: "Final_RMD"
author: "Matthew Brown"
date: "December 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(readr)
library(readxl)
library(stringr)
library(USAboundaries)
library(leaflet)
library(maps)
library(pander)


## Load in our centeroid data.
centeroids <- read_csv("Data/Idaho_Geo_County_Centeroids.csv")



# Gets the national data for every car crash involving a fatality.
get_accident_data <- function() {
  return(
    dat <- read_csv("Data/Accidents_Data.csv") %>%
      rename(COUNTY_NUMERIC = county) %>%
      mutate(STATE_ABV = state.abb[match(state_name,state.name)]) %>%
      mutate(SC_ID= paste(STATE_ABV, COUNTY_NUMERIC, sep='_')) %>%
      filter(state_name == "Idaho") %>%
      select( -COUNTY_NUMERIC)
  )
}


write_data <- function(dat) {
  dat %>%
    select(-geometry) %>%
    write.csv("Final_Data.csv")
}

# Gets the GLA data for county codes.
get_code_data <- function() {
  return(
    dat <- read_csv("Data/NationalFedCodes.csv") %>%
      select(COUNTY_NAME, STATE_ALPHA, COUNTY_NUMERIC) %>%
      group_by(COUNTY_NUMERIC, STATE_ALPHA) %>%
      filter(STATE_ALPHA == "ID") %>%
      filter(row_number(COUNTY_NAME) == 1) %>%
      ungroup() %>%
      mutate(SC_ID= paste(STATE_ALPHA, COUNTY_NUMERIC, sep='_')) %>%
      select(-STATE_ALPHA, -COUNTY_NUMERIC)
  )
}

# Returns a dataset containing the geometries of every county.
get_county_geometries <- function() {
  return (
    USAboundaries::us_counties() %>%
      filter(state_name == "Idaho") %>%
      filter(!is.na(name)) %>%
      mutate(SC_ID_NAMED = paste(state_name, name, sep='_')) %>%
      select(SC_ID_NAMED, geometry)
  )
}


# Returns a dataframe containing the 2010 population for each county.
get_county_populations <- function() {
  return (
    us_cities() %>%
      mutate(population, is.na(population), 0) %>%
      mutate(SC_ID_NAMED = paste(state_name, county_name, sep='_')) %>%
      group_by(SC_ID_NAMED) %>%
      summarize(county_pop = sum(population))
  )
}

# Merges the population and county geometry data.
merge_county_data <- function() {
  geometry_dat <- get_county_geometries()
  pop_dat <- get_county_populations()
  return(
    dplyr::left_join(geometry_dat, pop_dat, by="SC_ID_NAMED") %>%
      filter(!is.na(county_pop))
  )
}

merge_accident_code_dat <- function() {
  accident_dat <- get_accident_data() 
  code_dat <- get_code_data()
  return(
    dplyr::left_join(accident_dat, code_dat) %>%
      mutate(SC_ID_NAMED = paste(state_name, COUNTY_NAME, sep='_'))
  )
}

get_merged_data <- function() {
  accident_code_dat <- merge_accident_code_dat() 
  merged_county_dat <- merge_county_data()
  centeroids <- read_csv("Data/Idaho_Geo_County_Centeroids.csv") %>%
    mutate(county_latitude = as.numeric(county_latitude)) %>%
    mutate(county_longitude = as.numeric(county_longitude) * -1)
  
  accident_code_dat <- dplyr::left_join(centeroids, accident_code_dat)
   
  return(
    dplyr::left_join(accident_code_dat, merged_county_dat) %>%
      filter(!is.na(COUNTY_NAME))
  ) 
}

# Build our data
final_data <- get_merged_data()


```

```{r, echo=FALSE, warning=F}

```




```{r echo=FALSE, warning=FALSE}
## LEAFLET FUNCTIONS
display_crashes_by_state <- function(dat) {
  merged_data <- dat 
  
  ggplot()+
    geom_sf(data = merged_data, aes(fill = county_pop)) +
    scale_fill_gradient2()
}

display_Idaho_crashes <- function(dat) {
  merged_data <- dat %>%
    filter(STATE_ABV == "ID")
  ggplot()+
    geom_sf(data = merged_data, aes(fill = county_pop)) +
    scale_fill_gradient2()
    
}

display_fatalaties_by_county <- function(dat) {
  merged_data <- dat %>%
    filter(state_name == "California") %>%
    filter(county_pop >= 100000) %>%
    group_by(COUNTY_NAME) %>%
    mutate(total_fatalities = sum(number_of_fatalities)) %>%
    filter(row_number()==1) %>%
    mutate(total_fatalities = total_fatalities*100000/county_pop) %>%
    select(COUNTY_NAME, total_fatalities, county_pop, geometry) 
  
  ggplot()+
    geom_sf(data = merged_data, aes(fill = total_fatalities)) +
    scale_fill_gradient2()
}

leaflet_display_fatalaties_by_county <- function(dat) {

precise_accident_data <- final_data %>%
  mutate(county_pop = replace(county_pop, is.na(county_pop), 0)) %>%
  #filter(county_pop >= 0) %>%
  as_tibble()
  
  library(sf) 
  precise_accident_data <- precise_accident_data %>%
    st_as_sf()
  
  
  pal <- colorNumeric(c("white", "light blue", "pink"), 
                      min(precise_accident_data$county_pop):
                      max(precise_accident_data$county_pop)) 
  # pal <- colorNumeric(c("white", "black"), 
  #min(merged_data$total_fatalities):max(merged_data$total_fatalities)
  
  leaflet(precise_accident_data) %>%
    addTiles() %>%
    addPolygons(
      color = "#444444", weight = 1, smoothFactor = 0.5,
      opacity = 1, fillOpacity = 0.1,
      fillColor = ~pal(precise_accident_data$county_pop),
      highlightOptions = highlightOptions(color = "white", weight = 2,
      bringToFront = TRUE)
      ) %>%
    setView(lng=-112.033966, lat=43.491650, zoom=5) %>%
    addLegend("bottomright", pal = pal, values = ~county_pop,
              title = "County Population\n",
              opacity = 1
    ) %>%
    addCircles(lng=~longitude, 
               lat=~latitude, 
               weight=10,
               radius= ~1000*number_of_fatalities) %>%
    addLabelOnlyMarkers(
      ~county_longitude, 
      ~county_latitude - 0.1, 
      label= ~COUNTY_NAME, 
      labelOptions = 
        labelOptions(noHide = T, 
                     direction = 'top', 
                     textOnly = T))
    
}
## END LEAFLET FUNCTIONS
```

## Background
Car crashes in the United States results in over 30,000 fatalities annually. Kaggle.com hosts a dataset on every car related fatality for 2015. It is titled US Traffic Fatality Records. It contains many interesting features such as the light condition, if there was a drunk driver involved in the crash, and the atmospheric condition. Often times we have stereo types of what conditions lead to the safest driving experience. We often believe that poor atmospheric conditions, lighting, population of an area, and number of drunk drivers all contribute significantly to the number of car related fatalities. This analysis will focus only on one of these attributes. Specifically, it will focus on how the population of an area effects the fatality rate of an area. Is it possible to see if more densly populated areas increase the number of car crashes per capita? This analysis will look at traffic fatalities in Idaho.

## Analysis


## Population Analysis
```{r echo=FALSE, warning=F}
leaflet_display_fatalaties_by_county(final_data)
```


### County Populations

```{r echo=FALSE, warning=F}

idaho_county_populations <- final_data %>%
  group_by(COUNTY_NAME) %>%
  summarize(population = max(county_pop)) %>%
  mutate(population = replace(population, is.na(population), 0) / 1000) %>%
  ungroup() %>%
  as_tibble() 
  

idaho_county_populations$COUNTY_NAME <-
  factor(
    idaho_county_populations$COUNTY_NAME,
    levels = 
      idaho_county_populations$COUNTY_NAME[
        order(
          -idaho_county_populations$population
        )
        ]
  )

View(idaho_county_populations)

ggplot(
  data=idaho_county_populations,
  aes(x=COUNTY_NAME,y=population)
) +
  geom_col() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  ylab("Population (In Thousands)") +
  xlab("County Name") +
  ggtitle("Population Break Down of Idaho Counties (2010 Census)")
```

### Fatalities By County

```{r echo=FALSE, warning=F}


us_population_2010 <-  309300000
us_average_fatalities <- 31571 * 100000/ 309000000


idaho_fatality_county_rates <- final_data %>%
  group_by(COUNTY_NAME) %>%
  summarize(
    fatality_rate = sum(number_of_fatalities) * 100000 / max(county_pop),
    county_pop = max(county_pop)
    ) %>%
  ungroup() %>%
  arrange(desc(fatality_rate)) %>%
  mutate(highlighted = (county_pop >= 60000))

idaho_fatality_county_rates$COUNTY_NAME <-
  factor(
    idaho_fatality_county_rates$COUNTY_NAME,
    levels = 
      idaho_fatality_county_rates$COUNTY_NAME[
        order(
          -idaho_fatality_county_rates$fatality_rate
        )
        ]
  )


idaho_average_fatality_rate <- 15.0


################################################################
## BUILD THE IDAHO TOTAL FATALITIES DATA  
idaho_total_county_fatalities <- final_data %>%
  filter(state_name == "Idaho") %>%
  group_by(COUNTY_NAME) %>%
  summarize(total_fatalities = sum(number_of_fatalities),
            county_pop = max(county_pop)) %>%
  mutate(highlighted = case_when(
    county_pop >= 15000 ~ TRUE,
    county_pop < 15000 ~ FALSE,
    TRUE ~ FALSE
  ))



idaho_total_county_fatalities$COUNTY_NAME <-
  factor(
    idaho_total_county_fatalities$COUNTY_NAME,
    levels = 
      idaho_total_county_fatalities$COUNTY_NAME[
        order(
          -idaho_total_county_fatalities$total_fatalities
        )
        ]
  )

```

```{r echo=FALSE, warning=F}
# Displays the total number of fatalities.
ggplot() +
  geom_col(
    data=idaho_total_county_fatalities,
    aes(x=COUNTY_NAME, y=total_fatalities, fill=highlighted)
  ) +
  theme(axis.text.x = element_text(angle=45, hjust=1))  +
  scale_fill_discrete(guide = 'none') +
  geom_text(
    data=idaho_total_county_fatalities,
    aes(label = total_fatalities, y = total_fatalities + 0.1, x=COUNTY_NAME),
    position = position_dodge(0.9),
    vjust = 0
  ) +
  ylab("Total Fatalities") +
  xlab("County Name") +
  ggtitle("Total Number of Car Related Fatalities In Idaho For 2015") +
  scale_color_gradient(low = "white", high = "red")
```


## Fatality Rates Per 100,000 Peoples

```{r echo=FALSE, warning=F}

# Displays the total fatality rates.
ggplot() +
  geom_col(
    data=idaho_fatality_county_rates,
    aes(x=COUNTY_NAME, y=fatality_rate, fill=highlighted)
  ) +
  theme(axis.text.x = element_text(angle=45, hjust=1))  +
  geom_hline(aes(yintercept=us_average_fatalities)) +
  geom_text(
    aes(
      label="National Average Fatality Rate", 
      5, 
      us_average_fatalities, 
      vjust=-1)
  ) +
  geom_hline(aes(yintercept=idaho_average_fatality_rate)) +
  geom_text(
    aes(label="Idaho Average County Rate", 5,idaho_average_fatality_rate, vjust=-1)
  ) +
  scale_fill_discrete(guide = 'none') +
  ylab("Fatalities Per 100,000 People") +
  xlab("County Name") +
  ggtitle("Car Related Fatality Rates In Idaho For 2015") +
  geom_text(
    data=idaho_fatality_county_rates,
    aes(label = floor(fatality_rate), y = fatality_rate + 0.1, x=COUNTY_NAME),
    position = position_dodge(0.9),
    vjust = 0
  ) 

```

```{r echo=FALSE, warning=F}
  population_vs_fatalities <- final_data %>%
    filter(state_name=="Idaho") %>%
    filter(county_pop >= 0) %>%
    select(COUNTY_NAME, county_pop, number_of_fatalities) %>%
    group_by(COUNTY_NAME, county_pop) %>%
    summarize(
      total_fatalities = sum(number_of_fatalities)
    )
```

<br>

## Building The Dataset

The "Master Dataset", so to speak, contained five datasets that were merged together. Below, I describe each in detail.

<br>

#### US Fatalaties
Location: https://www.kaggle.com/usdot/nhtsa-traffic-fatalities <br>
To build this dataset, I had to use Python to interface with Google's Big Query. I selected the features below:

state_name <br>
county <br>
city <br>
hour_of_crash <br>
day_of_crash <br>
month_of_crash <br>
light_condition_name <br>
atmospheric_conditions_1_name <br>
number_of_fatalities <br>
number_of_drunk_drivers <br>
timestamp_of_crash <br>
latitude <br>
longitude <br>

<br>
The description for each feature can be found at the following link https://www.kaggle.com/usdot/nhtsa-traffic-fatalities.
<br>

This dataset needed to be cleaned. The biggest problem I had was that the counties feature did not have the actual name of the county. Instead, it had a government issued geo location code. In order to replace this code with a county name, I had to research what this code was. I then discovered a government-built dataset containing the names for each code. This is the National Fed Codes dataset discussed below. 

<br>

#### National Fed Codes
Location: https://www.gsa.gov/reference/geographic-locator-codes-glcs-overview <br>

This dataset contains the national geo location codes for each county in the United States. It's usefull if one wants to convert these codes into actual county names.

<br>

#### US County Geometries 
Location: USAboundaries R package <br>
I wanted to be able to display each county on my leaflet map. To do this, I merged the USAboundaries dataset with my accident dataset.

<br>

#### US Idaho County Populations
Location: https://www.idaho-demographics.com/counties_by_population
the USAboundaries datasets did not give me an accurate county population count. I had to find data on the populations of each county in Idaho. The above link has this data. I then used it to build a dataset containing the populations of each Idaho County.

<br>

#### County Geo Locations
Location: https://en.wikipedia.org/wiki/User:Michael_J/County_table <br>
This dataset contained the geometric centeroids of all the counties in Idaho. I gathered this data from Wikipedia and cleaned this data using regular expressions such as the one below: <br><br>
$\n[$^$+]*$
<br><br>
I used this dataset to place the names of each county marker on my leaflet map. It contains the center for each county in longitude and latitude coordinantes. 

<br>

#### Merging The Data
I merged all four datasets together. This provided me with a dataset that I then used to find the following for each county: The number of fatalities, the geographical centers, the population, and the geographical boundaries.